---
title: "R Basic Statistics Workshop"
author: "Melissa Ko"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
# we recommend using tidyverse for easier data processing
# but at a minimum you should install/load ggplot2 since it
# includes several useful built-in datasets!
# if you don't have tidyverse installed, comment out that line
library(tidyverse)
# install.packages("ggplot2")
library(ggplot2)
```

This handout is based on [prior work](https://darachm.github.io/dll-r/applying-basic-stats.html) by [Melissa Ko (mesako@stanford.edu)](https://github.com/mesako/) and Margaret with a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 License](https://github.com/darachm/dll-r/blob/main/LICENSE.md).

#### Instructions for this Document

Write your code and/or your response underneath each task prompt. To render this handout, click on the Knit button in Rstudio. You can also click on the green arrow in each code chunk to run the code that it contains.

------

#### This Workshop

This workshop aims to provide learners with the basic strategies to analyze a dataset in R, including:

1. describing quantitative variables
2. finding differences in data
3. identifying correlations
4. producing linear models

The goal is to enable learners to do their needed stats work in the R programming environment. **However, this is not an intro R workshop or a statistics class.** This workshop is recommended for learners who have some familiarity with R. Moreover, as a researcher, you know your data and the norms of your field best. Consider what statistics you must perform for your scenario and we will discuss how to do that in R!

------

#### Breakout Room Activity

During this breakout room activity, you will be randomly placed with a set of your peers. Please introduce yourselves! You may want to share your name, pronouns, department, and affiliation.

After everyone has introduced themselves, take turns sharing: how do you use statistics in your current research? What questions and/or approaches might come up in your work? Write your notes below.

*Your response here*

If you finish sharing early, please discuss your goals for what you want to learn to do in R through this workshop, and beyond.

------

##### Describing quantitative variables

```{r}
# msleep
# iris
```

```{r}
# hist()
```

**Task:** First create a histogram of a quantitative variable within either the `msleep` or `iris` dataset. You can choose to do this using the basic R `hist` function or apply your knowledge of using ggplot2.

```{r}
# Your code here
```

```{r}
# quantile()
# median()
# mean()
```

```{r}
# sd()
# IQR()
```

**Task:** Now calculate several measures to summarize the center, spread, and range of the quantitative variable you explored in the last task. Does what you see in the histogram match these statistics you calculated? 

```{r}
# Your code here
```

*Your response here*

------

##### Finding differences in data

```{r}
# t.test()
```

**Task:** Use the `t.test` function on the quantitative variable that you investigated in the prior section. Look at the confidence interval describing the uncertainty around the mean. How does this align with the histogram you created?

```{r}
# Your code here
```

*Your response here*

```{r}
# shapiro.test()
# var.test()
```

```{r}
# t.test(A, B, alternative = "less")
```

**Task:** Identify two different categorical groups present either in the `msleep` or `iris` data. Remember that you will be comparing the measurements of one group to the measurements of one other group. First create two new variables containing the measurements of each group separately. You may want to use the `filter` and `select` functions.

```{r}
# Your code here
```

**Task:** Now first determine if the assumptions for performing a t-test are true.

```{r}
# Your code here
```

*Your response here*

**Task:** Regardless of your results in the prior task, apply the `t.test` function to the data from the two groups. Interpret the results. 

```{r}
# Your code here
```

*Your response here*

```{r}
# wilcox.test()
```

```{r}
# aov()
# TukeyHSD()
```

**Task:** Apply the `aov` function to the `iris` dataset to determine whether a given quantitative variable differs significantly based on the `Species`. Describe your interpretation of these results.

```{r}
# Your code here
```

*Your response here*

**Task:** Apply the `aov` function to the `msleep` dataset to determine whether a given quantitative variable differs significantly based on the `conservation` status. Describe your interpretation of these results.

```{r}
# Your code here
```

*Your response here*

```{r}
# kruskal.test()
```

------

##### Identifying correlations

```{r}
# cor(A, B)
```

**Task:** Pick two quantitative variables in the `msleep` dataset or the `iris` dataset. Check that whether either of these columns have `NA` values. If needed, make a new tabular dataset where you remove the `NA` values, using functions like `drop_na`.

```{r}
# Your code here
```

**Task:** Using your dataset, create a scatterplot with a fitted trend line. Describe the plot and whether you observe a correlation between these two variables.

```{r}
# Your code here
```

*Your response here*

**Task:** Using the variable you created in the previous tasks, calculate both the Pearson or Spearman correlation using the `cor` function. Interpret these correlation values. 

```{r}
# Your code here
```

*Your response here*

------

##### Producing linear models

```{r}
# outcome ~ predictor1 + predictor2
# lm()
# summary(my.mod)
```

**Task:** Train a linear model using the `lm` function to predict one quantitative variable using another in the `iris` dataset. Look at the summary of this model and interpret the results.

```{r}
# Your code here
```

*Your response here*

**Task:** Train a linear model using the `lm` function to predict one quantitative variable using multiple quantitative variables in the `iris` dataset. Look at the summary of this model and interpret the results.

```{r}
# Your code here
```

*Your response here*

**Task:** For a challenge, train a linear model using the `lm` function to predict one quantitative variable using multiple variables including the `Species` column in the `iris` dataset. Look at the summary of this model and interpret the results.

```{r}
# Your code here
```

*Your response here*

```{r}
# plot(true, fit)
# predict()
```

------

#### Breakout Room Activity

During this breakout room activity, you will have the option to pick which of the following options you would like to work on with your peers.

The differences between these groups/options will be the dataset that you choose to apply statistics to: `midwest`, `diamonds`, or `infert`.

------

**Option 1: `midwest` Dataset** 

*Pick one of the following possible research questions:*

1. Is there a statistically significant difference in poverty rates between different states? What about a difference in percent college educated or professional?

2. Is there a statistically significant difference in any of these numerical measures between metro areas and non-metro areas?

3. Does one of the demographic features strongly correlate with percent educated? What about percent in poverty?

4. Do the measurements of the population demographics predict the level of poverty?

Based what you learned in this last section, describe some statistical approaches that may help you answer this question.

*Your response here*

Conduct the necessary statistical tests to draw any conclusions for this research question.

```{r}
# Your code here
```

------

**Option 2: `diamonds` Dataset**

*Pick one of the following possible research questions:*

1. Is there a statistically significant difference in cost between groups with different cuts? What about a difference between groups with different colors?

2. Does price strongly correlate with clarity? What about carat?

3. Do the measurements of the diamond (x, y, z) predict the price of the diamond?

Based what you learned in this last section, describe some statistical approaches that may help you answer this question.

*Your response here*

Conduct the necessary statistical tests to draw any conclusions for this research question.

```{r}
# Your code here
```

------

**Option 3: `infert` Dataset** 

*Pick one of the following possible research questions:*

1. Is there a statistically significant difference in parity between groups with different levels of education? What about a difference in abortions of either type?

2. Does age strongly correlate with parity? What about abortions of either type?

3. Does the number of prior induced abortions predict the number of prior spontaneous abortions?

Based what you learned in this last section, describe some statistical approaches that may help you answer this question.

*Your response here*

Conduct the necessary statistical tests to draw any conclusions for this research question.

```{r}
# Your code here
```
